Viscous Erosion in Porous Media

### 2018-02-22 docs/methods/erosion_methods.tex
NM

Because of the length and velocity scales of groundwater flow, we
restrict our attention to small Reynolds number flows.  For groundwater
flow, Darcy's law is often used, but this does not resolve the pore
structure and assumes a homogeneous geometry which is often not valid.
Instead, we use the incompressible Stokes equations to represent the
flow, and the individual pores are resolved which results in a complex
multiply-connected geometry.  A boundary integral equation is
well-suited for solving the incompressible Stokes equations in a complex
geometry since it can resolve individual grains and the non-negligible
interactions between bodies are resolved. Moreover, high-order accuracy
and computational efficiency are obtained. To link the eroding body and
fluid dynamics, the body is eroded in the normal direction at a rate
that is proportional to the magnitude of the shear
stress~\cite{ris-moo-chi-she-zha2012}.  Tracking the interface of
fluid-structure interaction problems with a long time horizon is
particularly---the mesh is often distorted and tangled by the dynamics
and regions of high curvature develop.  These two issues are resolved
by, first, introducing a tangential velocity field to maintain a grid
that is equispaced in arclength, but that does not affect the dynamics,
and, second, a curvature penalization term.  Furthermore, the frame of
reference is changed to \thL~variables so that the curvature term
appears as a diffusive term that can easily be treated implicitly
to avoid stiffness.

In addition to simulating the eroding grains, the flow is further
characterized by computing the vorticity in the flow.  The vorticity is
particularly insightful since it is identical to the shear stress on
no-slip boundaries in the Stokesian limit.  Therefore, by computing the
vorticity, we are able to locate regions where the erosion rates will be
fastest or slowest.

@@

To create high-fidelity and computationally efficient numerical methods,
two fast and accurate methods are employed.  First, the fluid equations
and the erosion rate (shear stress) are efficiently solved with spectral
accuracy by using a fast multipole accelerated boundary integral
equation method.  Given the rate of erosion, the shape is integrated in
time with second-order accuracy.  All spatial derivatives are be
computed in Fourier space, resulting in spectral accuracy in space.
Then, to further visualize the effect of erosion, we compute the
vorticity of the flow (Fig.~\ref{fig:50bodies}) to visualize regions
of high and low erosion.

@@

The second limitation is that the eroding bodies are fixed.  Mitchell
and Spagnolie~\cite{mit-spa2016} simulate a single mobile eroding body
in three dimensions~\cite{mit-spa2016}, but additional challenges are
caused when there are multiple bodies. In particular, especially in
dense suspensions, bodies can become extremely close to one another, and
the interaction between the bodies is difficult to resolve unless the
underlying mesh is refined to levels that are intractable. In future
work we plan to consider mobile eroding bodies, but include a small
force and torque so that unphysical contact is
avoided~\cite{lu-rah-zor2017}.

@@

We use a quasi-static approximation.  That is, at each time step, we
freeze the geometry, solve the incompressible Stokes equations, and then
update the geometry according to the erosion model.  The two steps
inside the red rectangle describes a time step in the quasi-static
approximation.
%A flow diagram is illustrated in Figure~\ref{fig:workflow}.  
%\begin{figure}[htpb]
%  \centering
%  \input{figs/workflow.tikz}
%  \caption{\label{fig:workflow}A caption.}
%\end{figure}

Our erosion model is based on the shear stress, and this model is
discussed in more detail in Section~\ref{sec:erosion}.  We then describe
how the incompressible Stokes equations are solved in $\Omega$ using a
boundary integral equation in Section~\ref{sec:bies}.  In
Section~\ref{sec:shearStressLP}, we discuss how the shear stress is
computed.  Finally, in Section~\ref{sec:thetaL}, we describe how corners
in the geometry are avoided and perform a change of coordinates that
linearizes a stiff term.

>>>>

Because of the length and velocity scales of groundwater flow, we
restrict our attention to small Reynolds number flows.  For groundwater
flow, Darcy's law is often used, but this does not resolve the pore
structure and assumes a homogeneous geometry which is often not valid.
Instead, we use the incompressible Stokes equations to represent the
flow, and the individual pores are resolved which results in a complex
multiply-connected geometry.  A boundary integral equation is
well-suited for solving the incompressible Stokes equations in a complex
geometry since it can resolve individual grains and the non-negligible
interactions between bodies are resolved. Moreover, high-order accuracy
and computational efficiency are obtained. To link the eroding body and
fluid dynamics, the body is eroded in the normal direction at a rate
that is proportional to the magnitude of the shear
stress~\cite{ris-moo-chi-she-zha2012}.  Tracking the interface of
fluid-structure interaction problems with a long time horizon is
particularly---the mesh is often distorted and tangled by the dynamics
and regions of high curvature develop.  These two issues are resolved
by, first, introducing a tangential velocity field to maintain a grid
that is equispaced in arc length, but that does not affect the dynamics,
and, second, a curvature penalization term.  Furthermore, the frame of
reference is changed to \thL~variables so that the curvature term
appears as a diffusive term that can easily be treated implicitly
to avoid stiffness.

In addition to simulating the eroding grains, the flow is further
characterized by computing the vorticity in the flow.  The vorticity is
particularly insightful since it is identical to the shear stress on
no-slip boundaries.  Therefore, by computing the
vorticity, we are able to locate regions where the erosion rates will be
fastest or slowest.

\todo[inline]{Note: I took out 'in the Stokesian limit' because the vorticity matches the shear stress on the boundary no matter what the Reynolds number is.}

@@

To create high-fidelity and computationally efficient numerical methods,
two fast and accurate methods are employed.  First, the fluid equations
and the erosion rate (shear stress) are efficiently solved with spectral
accuracy by using a fast-multipole-accelerated boundary integral
equation method.  Given the rate of erosion, the shape is integrated in
time with second-order accuracy.  All spatial derivatives are be
computed in Fourier space, resulting in spectral accuracy in space.
To further visualize the effect of erosion, we compute the
vorticity of the flow (Fig.~\ref{fig:50bodies}) to visualize regions
of high and low erosion.

@@

The second limitation is that the eroding bodies are fixed.  Mitchell
and Spagnolie~\cite{mit-spa2016} simulate a single mobile eroding body
in three dimensions~\cite{mit-spa2016}, but additional challenges are
caused when there are multiple bodies. In particular, especially in
dense suspensions, bodies can become extremely close to one another, and
the interaction between the bodies is difficult to resolve unless the
underlying mesh is refined to levels that are intractable. In future
work we plan to consider mobile eroding bodies by including a small
force and torque to prevent unphysical contact~\cite{lu-rah-zor2017}.

@@

We use a quasi-static approximation.  That is, at each time step, we
freeze the geometry, solve the incompressible Stokes equations, and then
update the geometry according to the erosion model.  The two steps
inside the red rectangle describes a time step in the quasi-static
approximation.
\todo[inline]{Appeal to separation of scales to justify this quasi-static approximation; i.e.~the timescale of erosion is much longer than that of the flow.}

%A flow diagram is illustrated in Figure~\ref{fig:workflow}.  
%\begin{figure}[htpb]
%  \centering
%  \input{figs/workflow.tikz}
%  \caption{\label{fig:workflow}A caption.}
%\end{figure}

Our erosion model is based on the shear stress, and this model is
discussed in more detail in Section~\ref{sec:erosion}.  We then describe
how the incompressible Stokes equations are solved in $\Omega$ using a
boundary integral equation in Section~\ref{sec:bies}.  In
Section~\ref{sec:shearStressLP}, we discuss how the shear stress is
computed.  Finally, in Section~\ref{sec:thetaL}, we describe how corners
in the geometry are avoided and perform a change of coordinates to treat a stiff term.
\todo[inline]{I took out all references to the stiff term being linear, because with the new scaling we use (computing the mean shear stress on each body) the diffusive term is actually nonlinear (because the mean shear stress depends on theta). Nonetheless, the integrating-factor method still works.}
